---
title: NHL API - Collect Rosters
author: ' Nick Paul'
date: '2021-12-07'
slug: []
categories: ['Data-Collection']
tags: ['python', 'NHL']
subtitle: ''
description: 'In this post we will explore collecting NHL roster data.'
image: ''
---

1. [Import Libraries](#import-libs)
1. [Collect Team Info](#collect-team-info)
1. [Collect Roster](#collect-roster)
1. [Collect Career Stats](#collect-career-stats)
1. [Collect All Players in a Season](#collect-all-players-in-a-season)

In this post we will explore how to get roster data for each NHL team. We will then look at how to get career stats for each player on a particular roster. Finally, we will get career stats for all players from a particular season.

```{r message=FALSE, warning=FALSE, include=FALSE}
reticulate::use_python("/home/rstudio/.local/share/r-miniconda/envs/pandas/bin/python",required = TRUE)
```


```{r include=FALSE}
library(reticulate)
use_condaenv("pandas")
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE
)
```

## Import Libraries

```{python}
import requests
import json
import pandas as pd
```

## Collect Team Info

The first step is to collect data about NHL teams. This will allow us to focus on active NHL teams.

```{python}

teams_url = "https://statsapi.web.nhl.com/api/v1/teams"
team_response = requests.get(teams_url)

team_content = json.loads(team_response.content)
df_teams = pd.json_normalize(team_content['teams'],
sep = "_")

df_teams.info()
```

## Collect Roster

Next, we create a function to get a team roster for a particular season.


```{python}
df_teams['link'] = 'https://statsapi.web.nhl.com' + df_teams['link']
df_active = df_teams.loc[df_teams['active']==True]

def get_team_roster(team, season):

    base_url = df_active.loc[df_active['name']==team]['link'].iloc[0]
    print(base_url)
    url = base_url + "/roster/" + "?season=" + season
    
    response = requests.get(url)
    roster = response.json()["roster"]
    
    df_roster = pd.json_normalize(roster, sep = "_").astype(str)

    return df_roster
    
df_roster = get_team_roster("New York Rangers", '20192020')
df_roster
```

## Collect Career Stats

Now that we have player IDs for each player on a team, lets create a function that gets year-by-year stats for a particular player.

```{python}
from datetime import datetime
import numpy as np

def get_career_stats(player_id):

      
    url = 'https://statsapi.web.nhl.com/api/v1/people/' + player_id + '/stats/?stats=yearByYear'
    response = requests.get(url)
    content = json.loads(response.content)['stats']
    splits = content[0]['splits']

    df_splits = (pd.json_normalize(splits, sep = "_" )
             .query('league_name == "National Hockey League"')
            )
    if df_splits.shape[0] > 0 :
    
        url_info = 'https://statsapi.web.nhl.com/api/v1/people/' + player_id
        response = requests.get(url_info)
        player_info = json.loads(response.content)['people'][0]

        if player_info['primaryPosition']['code'] != "G":
            df_splits['goals_per_game']=  df_splits['stat_goals']/df_splits['stat_games']
        df_splits['player_id'] = player_id
        df_splits['first_name'] = player_info['firstName']
        df_splits['last_name'] = player_info['lastName']
        df_splits['bday'] = pd.to_datetime(player_info['birthDate'])
        df_splits['season_end_yr'] = [x[4:8] for x in df_splits['season']]
        df_splits['season_start_yr'] = [x[0:4] for x in df_splits['season']]
        df_splits['season_start_dt'] =  [datetime.strptime(x + '0930', "%Y%m%d") for x in df_splits['season_start_yr']] 
        df_splits['age'] = (np.floor((df_splits['season_start_dt'] - df_splits['bday'])/ np.timedelta64(1,'Y') ))
        df_splits['age'] = df_splits['age'].astype(int)
    
    return df_splits
    
get_career_stats(player_id = '8478402')
```




Now we can loop over the player ids from a roster and collect year-by-year stats for each player.

```{python}
stats = []
for player_id in df_roster['person_id']:
    df = get_career_stats(player_id)
    stats.append(df)
df_all = pd.concat(stats)
df_all
```

## Collect All Players in a Season

Next, we can loop over every team and collect rosters for all active teams from a particular seaason.

```{python}
rosters = []
season = '20212022'
for team in df_active['name']:
    df_roster = get_team_roster(team, season)
    rosters.append(df_roster)

df_all_rosters = pd.concat(rosters)
df_all_rosters
```

Finally, we can loop over all players to get year-by-year stats for all players during a particular seasn.

```{python, eval = FALSE}
stats = []
for player_id in df_all_rosters['person_id']:
    print(player_id)
    df = get_career_stats(player_id)
    stats.append(df)

df_all_stats = pd.concat(stats)
df_all_stats
```



```{python, eval = FALSE}
df_all_stats.to_csv("nhl_20212022_career_stats.csv")
```

