---
title: NHL API - Collect Career Stats 
author: ' Nick Paul'
date: '2021-12-03'
slug: []
categories: ['Data-Collection']
tags: ['python', 'NHL', 'API']
subtitle: ''
description: 'This post demonstrates how to collect career stats for an NHL player'
image: ''
---

1. [Import Libraries](#import-libs)
1. [Find Player ID](#find-player-id)
1. [Collect Data](#collect-data)


```{r message=FALSE, warning=FALSE, include=FALSE}
reticulate::use_python("/home/rstudio/.local/share/r-miniconda/envs/pandas/bin/python",required = TRUE)
```


```{r include=FALSE}
library(reticulate)
use_condaenv("pandas")
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE
)
```

The NHL has an unpublished API that makes grabbing stats pretty painless.  In this post we will walk through how to grab year-by-year stats for a particular player using python.

## Import Libraries

To get started, we will import the following libraries:

* `requests` - Conduct the API call
* `json` - Convert the returned data to a python list
* `pandas` - Conduct data manipulation 

```{python}
import requests
import json
import pandas as pd
```

## Find Player ID

To collect player data from the NHL API, we first need their player ID, which can be found using the NHL's suggest end point.  Below is a simple example to get `Connor McDavid's` player ID.

```{python}
base_url = 'https://suggest.svc.nhl.com/svc/suggest/v1/minplayers/'

first_name = 'connor'
last_name = 'mcdavid'
num_to_return = '1'

full_url = base_url + first_name + '%20' + last_name + '/' + num_to_return
response = requests.get(full_url)
suggestions = json.loads(response.content)['suggestions'][0]
print(suggestions)
player_info = str.split(suggestions, "|")
print(player_info)
player_id = player_info[0]
player_id
```

This endpoint returns a string that we can split and pull out information such as name and birthday. 

## Career Data

The next step is to hit the stats api using the player ID we just acquired.


```{python}
url = 'https://statsapi.web.nhl.com/api/v1/people/' + player_id + '/stats/?stats=yearByYear'
response = requests.get(url)
content = json.loads(response.content)['stats']
splits = content[0]['splits']

df_splits = (pd.json_normalize(splits, sep = "_" )
             .query('league_name == "National Hockey League"')
            )
df_splits
```

We now have Connor McDavid's career stats! While McDavid's stats returned one row per year, be aware that a player that was traded during a season will have multiple rows for a particular year.

## Create a Function

Since we will likely reuse this code, we should write a function that handles the two previous steps: get a player ID and get career stats. 

```{python}
from datetime import datetime
import numpy as np

def get_career_stats(first_name, last_name):

    base_url = 'https://suggest.svc.nhl.com/svc/suggest/v1/minplayers/'

    num_to_return = '1'

    full_url = base_url + first_name + '%20' + last_name + '/' + num_to_return

    response = requests.get(full_url)
    suggestion = json.loads(response.content)['suggestions'][0]
    player_info = str.split(suggestion, "|")
    player_id = player_info[0]
    
    url = 'https://statsapi.web.nhl.com/api/v1/people/' + player_id + '/stats/?stats=yearByYear'
    response = requests.get(url)
    content = json.loads(response.content)['stats']
    splits = content[0]['splits']

    df_splits = (pd.json_normalize(splits, sep = "_" )
             .query('league_name == "National Hockey League"')
            )
    
    df_splits['goals_per_game']=  df_splits['stat_goals']/df_splits['stat_games']
    df_splits['player_id'] = player_id
    df_splits['first_name'] = player_info[2]
    df_splits['last_name'] = player_info[1]
    df_splits['bday'] = pd.to_datetime(player_info[10])
    df_splits['season_end'] = [x[4:8] for x in df_splits['season']]
    df_splits['season_start_yr'] = [x[0:4] for x in df_splits['season']]
    df_splits['season_start_dt'] =  [datetime.strptime(x + '0930', "%Y%m%d") for x in df_splits['season_start_yr']] 
    df_splits['age'] = (np.floor((df_splits['season_start_dt'] - df_splits['bday'])/ np.timedelta64(1,'Y') ))
    df_splits['age'] = df_splits['age'].astype(int)
    
    return df_splits
get_career_stats('connor', 'mcdavid')
```

Now we can just enter a players name and pull their stats, which is very useful if we want to pull data for multiple players and compare their numbers.

```{python}
df_ovi = get_career_stats('connor', 'mcdavid')
df_mcdavid = get_career_stats('alex', 'ovech')

df_compare = pd.concat([df_ovi, df_mcdavid])

df_compare.groupby(['first_name', 'last_name'])['stat_goals'].sum()
```

